New approaches to teachers' experience of stress: Do heart rate measurements with fitness trackers provide an efficient, inexpensive, and robust measurement method?

### 2022-11-25 02_paper/02_study/02_fitbit paper/fitbit_paper.Rmd
fitbit paper participants

# demographic data
df_demo <- 
  read_excel("./data/data_empschul_labor_lehrperson.xlsx") %>% # read in excel 
  transmute(ID = LI06_05, # select and rename relevant columns 
            # gender = factor(LI02_01_1,
            #                 levels = 1:2,
            #                 labels = c("male","female")
            #                 ),
            gender = LI02_01_1, # 1 = male; 2 = female
            age = LI03_01, 
            teaching_experience = LI04_01)
            # school_type = LI18)
            # term = LI11_01, 
            # extracurricular_teaching_experience = LI14,
            # secondary_teaching_activities = LI17)

# merge two data frames by ID 
df_merge <- merge(df_demo,
                  df_rating,
                  by = "ID") %>% 
  filter(!ID %in% c("126", # exclude cases with no fitbit data (to check see data Heart Rate)
                    "132")
         )

@@

@@

# create a basic table (tibble) using tidyverse functions
df_demo_table <- df_demo %>%
  summarise(N=n(),
            "Gender female in percent" = round(sum(gender/sum_participants) * 100,
                                               digits = 2),
            "M Age in years" = round(mean(age),
                                     digits = 2),
            "SD Age in years" = round(sd(age),
                                      digits = 2),
            "Min Age in years" = min(age),
            "Max Age in years" = max(age),
            "M Teaching Experience in years" = round(mean(teaching_experience),
                                            digits = 2),
            "SD Teaching Experience in years" = round(sd(teaching_experience),
                                            digits = 2),
            "Min Teaching Experience in years" = min(teaching_experience),
            "Max Teaching Experience in years" = max(teaching_experience)
            )

@@

The sample consisted of *N* = `r length(df_demo$ID)` participants. Data of two participants were excluded from the analyses due to insufficient data quality.

@@

## Material
We used a Fitbit Charge 4 to measure the teachers' heart rate. The device was attached 2-finger widths above the ulnar styloid process to the subject’s wrist following the manufacturer’s instructions. To determine the heart rate, the Fitbit flashes green LEDs many times per second and uses light-sensitive photodiodes to measure the volume changes in the capillaries and then calculates how many times the heart beats per minute (bpm). Data were automatically wireless synced with an iPad via Bluetooth to a Fitbit account, and subsequently, the intraday second-by-second data were exported for each session using the opensource software Pulse Watch (PulseWatch. URL: https://iccir919.github.io/pulseWatch/public/index.html [accessed 2022-08-03]).

## Procedure
In a laboratory setting, three trained actors performed teaching disruptions in a counter balanced fashion while the subject taught a 15-minute micro-teaching unit prepared in advance. To record the subject's heart rate, the Fitbit Charge 4 was put on at least 10 minutes before the start of the lesson. The lesson was recorded by four cameras and an audio recorder. In addition, the subject wore eye-tracking glasses to record gaze behavior.

@@

The Fitbit watch was removed only after the last questionnaire to obtain heart rate at three different measurement time points: before, during, and after the lesson.

@@

The heart rate of each subject were measured before, during and after the experiment.

```{r}
# Initial Preparations

## Run following Code only if new data is added to the initial data.
# 
# source(file = "data.wrangling.R")
# 
# # aggregate data 
# df_heart_all <-
#   read_rds(file = "data/heart_all.rds") |> 
#   mutate(status = factor(status,
#                          levels = c("Expert","Novice"),
#                          labels = c("Expert","Novice")
#                          )
#          )
# 
# df_heart_min <-
#   read_rds(file = "data/heart_min.rds")
# 
# df_heart_aggr <- 
#   read_rds(file = "data/heart_aggr.rds") |> 
#   mutate(status = factor(status,
#                          levels = c("Novice","Expert"),
#                          labels = c("Novice","Expert")
#                          )
#          )
# 
# 
# # Mean, SD, t-test & Cohen D for experts & novices
# # novices
# mean_nov <- 
#   df_heart_aggr %>%
#   filter(status == "Novice" ) %>% 
#   pull(mean) %>% 
#   mean() %>% 
#   round(., digits = 0)
# 
# sd_nov <- 
#   df_heart_aggr %>%
#   filter(status == "Novice") %>% 
#   pull(mean) %>% 
#   sd() %>% 
#   round(., digits = 0)
# 
# # experts
# mean_exp <-
#   df_heart_aggr %>%
#   filter(status == "Expert") %>% 
#   pull(mean) %>% 
#   mean() %>% 
#   round(., digits = 0)
# 
# sd_exp <- 
#   df_heart_aggr %>%
#   filter(status == "Expert") %>% 
#   pull(mean) %>% 
#   sd() %>% 
#   round(., digits = 0)
# 
# # t-test for expertise differences
# t.test(x = df_heart_aggr$mean[df_heart_aggr$status == "Expert"],
#        y = df_heart_aggr$mean[df_heart_aggr$status == "Novice"])
# 
# 
# # effect size for expertise differences
# d_heartrate <- CohenD(x = df_heart_aggr$mean[df_heart_aggr$status == "Expert"],
#                y = df_heart_aggr$mean[df_heart_aggr$status == "Novice"],
#                na.rm = TRUE)
# 
# 
# # Plots
# 
# plot_flow_leg <-
# ggplot(data = df_heart_all,
#        mapping = aes(x = Time,
#                      y = `Heart Rate`
#                      )
#        ) +
#   xlim(0, 600) +
#   geom_smooth(mapping = aes(color = status),
#               method = "loess") +
#   scale_color_viridis_d(option = "A",
#                         end = 0.8) +
#   scale_linetype_manual(values = c(1, 6)) +
#   labs(x = "Time (in Seconds)",
#        y = "Heart Rate\n(in Beats per Minute)") +
#   ggtitle("Experts' and novices' heart rate per phase") +
#   theme_minimal() +
#   theme(text = element_text(family = "serif"),
#         legend.text = element_text(size = 15),
#         legend.title = element_blank(),
#         axis.text.x = element_text(size = 12),
#         axis.text.y = element_text(size = 10),
#         strip.text = element_text(size = 16),
#         plot.title = element_text(size = 20, 
#                                   face = "bold")) +
#   facet_grid(cols = vars(Type),
#              scales = "free_x")
# 
# # saving plot 
# ggsave(plot = plot_flow_leg,
#        filename = "plots/plot_flow_leg.svg",
#        height = 5,
#        width = 8,
#        units = "in")

### Steps

Using the FitBit we measured the steps a person walked during the experiment as well. We validated the steps in six randomly selected videos by manual step counting. A step was considered to be the movement of the subject's foot forward, backward, or sideways. The measurement was not as reliable as hoped.

```{r}
# df_steps <- read_rds(file = "data/steps.rds")
# set.seed(5678)
# ggplot(data = df_steps,
#        mapping = aes(x = type,
#                      y = Steps
#                      )
#        ) +
#   geom_boxplot(mapping = aes(color = type),
#                fill = "grey",
#                alpha = 0.6) +
#   geom_jitter(mapping = aes(color = type),
#               alpha = 0.9,
#               width = 0.3) +
#   scale_color_viridis_d(option = "A",
#                        end = 0.8) +
#   labs(x = "",
#        y = "Steps During\nMain Phase") +
#   theme_minimal() +
#   theme(text = element_text(family = "serif"),
#         legend.title = element_blank(),
#         axis.text.x = element_blank(),
#         axis.ticks.x = element_blank()
#         )
# 
# ### Validation of steps
# # selecting some rows at random with sample_n()
# sample_n(df_steps, 6)
# 
# # random ID Output:
# # Novices ID: 102, 125, 110
# # Experts ID: 219, 218, 214
# 
# # Results from manual step counting:
# # Novices ID: 110 -> 178 steps; 102 -> 219 steps; 125 -> 319 steps
# # Experts ID: 214 -> 437 steps; 218 -> 195 steps; 219 -> 530 steps

>>>>

# demographic data
df_demo_all <- 
  read_excel("./data/data_empschul_labor_lehrperson.xlsx") %>% # read in excel 
  transmute(ID = LI06_05, # select and rename relevant columns 
            # gender = factor(LI02_01_1,
            #                 levels = 1:2,
            #                 labels = c("male","female")
            #                 ),
            gender = LI02_01_1, # 1 = male; 2 = female
            age = LI03_01, 
            teaching_experience = LI04_01,
            school_type = LI18,
            term = LI11_01,
            extracurricular_teaching_experience = LI14,
            secondary_teaching_activities = LI17)

# merge two data frames by ID 
df_merge <- merge(df_demo_all,
                  df_rating,
                  by = "ID") %>% 
  filter(!ID %in% c("126", # exclude cases with no fitbit data (to check see data Heart Rate)
                    "132")
         )

@@

# exclude cases with no fitbit data
df_demo <- df_demo_all %>% 
  filter(!ID %in% c("126",
                     "132"
                     )
          )

@@

# replacing 2 to 1 (Oberschule = Gymnasisium --> Secondary school)
df_demo$school_type[df_demo$school_type == 2] <- "1"

# creating new column to calculate percent of school type
df_demo <- 
  df_demo %>%
  group_by(school_type) %>%
  mutate(count_schooltype = sum(n())) %>% 
  mutate(percent_schooltype = round((count_schooltype/sum_participants) * 100, 
         digits = 2))

# changing character Percent_schooltype into numeric
df_demo$percent_schooltype <- as.numeric(df_demo$percent_schooltype)

#changing type of school into character
# LI18:
# 1: Grundschule
# 2: Oberschule
# 3: Gymnasium
# 4: Förderschule/Sondershausen
# 5: Berufsbildende Schule
# -1: andere Schulform
df_demo$school_type <- factor(x = df_demo$school_type,
                              levels = c("1","3","4","5","-1"), # Mögliche Werte # andere Möglichkeit: unique(demo.data$School_type)
                              labels = c("Primary School","Secondary School", "Special Education","Vocational School", "Other Schooltype")) # Bezeichnungen

# create a basic table (tibble) using tidyverse functions
df_demo_table <- df_demo %>%
  ungroup() %>%
  summarise(N=n(),
            "Gender female in percent" = round(sum(gender/sum_participants) * 100,
                                               digits = 2),
            "M Age in years" = round(mean(age),
                                     digits = 2),
            "SD Age in years" = round(sd(age),
                                      digits = 2),
            "Min Age in years" = min(age),
            "Max Age in years" = max(age),
            "M Teaching Experience in years" = round(mean(teaching_experience),
                                            digits = 2),
            "SD Teaching Experience in years" = round(sd(teaching_experience),
                                            digits = 2),
            "Min Teaching Experience in years" = min(teaching_experience),
            "Max Teaching Experience in years" = max(teaching_experience)
            )

@@

The sample consisted in total of *N* = `r length(df_demo_all$ID)` pre- and in-service teachers. The subjects were recruited from the Leipzig University or from German schools in Saxony via personal contact, e-mail lists and flyers. Data of two participants were excluded from the analyses due to insufficient data quality, yielding demographic data from *N* = `r length(df_demo$ID)` subjects.

@@

`r df_demo %>% filter(school_type == "Primary School") %>% pull("percent_schooltype") %>% unique()`% of the subjects were (studying to become) teachers for primary school, `r df_demo %>% filter(school_type == "Secondary School") %>% pull("percent_schooltype") %>% unique()`% were (studying to become) teachers for secondary school and  `r df_demo %>% filter(school_type == "Special Education") %>% pull("percent_schooltype") %>% unique()`% were (studying to become) teachers for special education needs.

All study procedures were carried out in accordance with the ethical standards of the University’s Institutional Review Board and the authors received a positive vote on the study procedures from the Ethics Committee Board of Leipzig University. All participants were informed in detail about the aim and intention of the study prior to testing. Participation in the study was voluntary and only took place after written consent has been given.

## Material

### Teachers' heart rate
We used a Fitbit Charge 4 to measure the teachers' heart rate. The device was attached 2-finger widths above the ulnar styloid process to the subject’s wrist following the manufacturer’s instructions. To determine the heart rate, the Fitbit flashes green LEDs many times per second and uses light-sensitive photodiodes to measure the volume changes in the capillaries and then calculates how many times the heart beats per minute (bpm). Data were automatically wireless synced with an iPad via Bluetooth to a Fitbit account, and subsequently, the intraday second-by-second data were exported for each session using the opensource software Pulse Watch (PulseWatch. URL: https://iccir919.github.io/pulseWatch/public/index.html [accessed 2022-08-03]).

### Teachers' self-assessment of arousal during the teaching phase

## Procedure

The data collection was part of a larger research project with a planned sample of *n* = 40 in-service teachers and *n* = 40 pre-service teachers.

In a controlled laboratory setting, heart rate data in beats per minute (bpm) were recorded using Fitbit Charge 4 over a total period of approximately two hours. Within this time frame, teachers taught a 15-minute self-prepared lesson to an audience of three actors. For analyzing the heart rate data, we selected five 10-minute intervals of theoretical interest: 
(1) the pre-teaching phase, which started from the moment the subject put on the watch 
(2) the teaching phase, which began at the moment when the subject started the lesson 
(3) the post-teaching phase, which was the 10 minutes immediately after the end of the lesson
(4) the interview phase, which was 10 minutes in the middle of the interview (for this, the difference was calculated from the end of the lesson and from the time when the subject took off the watch. This duration was divided in two to get to the middle of the interval and 5 minutes were subtracted to get to the start of the 10-minute interval),
(5) the end phase, which was the 10 minutes before the subject took off the watch.

In a laboratory setting, three trained actors performed teaching disruptions in a counter balanced fashion while the subject taught a 15-minute micro-teaching unit prepared in advance. To record the subject's heart rate, the Fitbit Charge 4 was put on at least 10 minutes before the start of the lesson. The lesson was recorded by four cameras and an audio recorder. In addition, the subject wore eye-tracking glasses to record gaze behavior.

@@

The Fitbit watch was removed only after the last questionnaire to obtain heart rate data during the entire study.

@@

### 2022-11-25 02_paper/02_study/02_fitbit paper/fitbit_paper.tex
fitbit paper participants

\begin{table}[h]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:demographicstable}Demographic Information}

\tiny{

\begin{tabular}{llllllllll}
\toprule
N & \multicolumn{1}{c}{Gender female in percent} & \multicolumn{1}{c}{M Age in years} & \multicolumn{1}{c}{SD Age in years} & \multicolumn{1}{c}{Min Age in years} & \multicolumn{1}{c}{Max Age in years} & \multicolumn{1}{c}{M Teaching Experience in years} & \multicolumn{1}{c}{SD Teaching Experience in years} & \multicolumn{1}{c}{Min Teaching Experience in years} & \multicolumn{1}{c}{Max Teaching Experience in years}\\
\midrule
63 & 63.49 & 29.32 & 10.34 & 19.00 & 59.00 & 4.51 & 9.18 & 0.00 & 37.00\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

The sample consisted of \emph{N} = 63 participants. Data of two participants were excluded from the analyses due to insufficient data quality.

The subjects (40 women; 63.49 \%) had a mean age of 29.30 years (\emph{SD} = 10.30; range: 19-59) and an average teaching experience of 4.51 years (\emph{SD} = 9.18; range: 0-37).

@@

@@

>>>>

The sample consisted in total of \emph{N} = 63 pre- and in-service teachers. The subjects were recruited from the Leipzig University or from German schools in Saxony via personal contact, e-mail lists and flyers. Data of two participants were excluded from the analyses due to insufficient data quality, yielding demographic data from \emph{N} = 61 subjects.

The subjects (39 women; 63.93 \%) had a mean age of 29.60 years (\emph{SD} = 10.40; range: 19-59) and an average teaching experience of 4.66 years (\emph{SD} = 9.30; range: 0-37).

18.03\% of the subjects were (studying to become) teachers for primary school, 72.13\% were (studying to become) teachers for secondary school and 9.84\% were (studying to become) teachers for special education needs.

All study procedures were carried out in accordance with the ethical standards of the University's Institutional Review Board and the authors received a positive vote on the study procedures from the Ethics Committee Board of Leipzig University. All participants were informed in detail about the aim and intention of the study prior to testing. Participation in the study was voluntary and only took place after written consent has been given.

@@

\hypertarget{teachers-heart-rate}{%
\subsubsection{Teachers' heart rate}\label{teachers-heart-rate}}

@@

\hypertarget{teachers-self-assessment-of-arousal-during-the-teaching-phase}{%
\subsubsection{Teachers' self-assessment of arousal during the teaching phase}\label{teachers-self-assessment-of-arousal-during-the-teaching-phase}}
